---
title: "Phage-Host Prediction Analysis"
params:
  name: "02-phage-host-prediction"
---

# Phage-Host Prediction for Animal Pathogens

## Setup

```{r setup}
here::i_am(paste0(params$name, ".qmd"), uuid = "b2c3d4e5-f6a7-8901-bcde-f1234567890")

# qproj setup
qproj::proj_create_dir_target(params$name)
path_target <- qproj::path_target(params$name)
path_source <- qproj::path_source(params$name)

# Read upstream data
path_upstream <- qproj::path_source("01-public-data-mining")
```

## Load Packages

```{r packages}
library(tidyverse)
library(here)
library(qproj)
library(ape)
library(ggplot2)
# For phage-host prediction
# library(phageR)      # If available
# library(vConTACT2R)  # If available
```

## 1. Objective

Predict which phages can effectively target specific animal pathogens:
- **Target pathogens**: *Clostridium perfringens*, *Salmonella enterica*, *E. coli*, *Vibrio* spp.
- **Goal**: Identify lytic phages with broad host range or specific activity
- **Approach**: Use computational tools + existing databases (phagesDB, MVP, vConTACT3)

## 2. Phage-Host Prediction Tools Review

Based on NotebookLM research (34 sources imported), key tools include:

### 2.1 Sequence-Based Methods
- **vConTACT3**: Viral protein cluster networks, genus-level prediction
- **PHIAF**: Phage-host interaction predictor using k-mer features
- **DeepHost**: Deep learning approach using genomic features
- **WIsH**: k-mer based method

### 2.2 Structure-Based Methods
- **PHP**: Prokaryotic host prediction using structural features
- **VHQS**: Quality scoring for predicted interactions

### 2.3 Integrated Pipelines
- **ViroProfiler** (our existing tool)
- **TPD** (Therapeutic Phage Discovery pipeline)

## 3. Database Integration

### 3.1 Download Phage Genomes from RefSeq

Template for downloading phage genomes targeting our pathogens of interest.

```{r download-phage-genomes}
#| eval: false

# Using entrez to download phage genomes
if (FALSE) {  # Set to TRUE when ready
  library(rentrez)
  
  # Search for phages targeting C. perfringens
  search_cperf <- rentrez::entrez_search(
    db = "nucleotide",
    term = "Clostridium perfringens phage[Organism] AND refseq[filter]",
    retmax = 100
  )
  
  # Download sequences
  phage_seqs <- rentrez::entrez_fetch(
    db = "nucleotide",
    id = search_cperf$ids,
    rettype = "fasta"
  )
  
  # Save
  writeLines(
    phage_seqs,
    file.path(path_target, "raw", "c_perfringens_phages.fasta")
  )
}
```

### 3.2 Load Existing Database

Load phage-host interaction data from integrated databases.

```{r load-databases}
#| eval: false

# Template: Load phagesDB data
# Source: https://phagesdb.org/download/
phagesdb_url <- "https://phagesdb.org/media/ExportDatasets/"

# Template: Load vConTACT3 results if available
# vcontact_results <- read_tsv(file.path(path_upstream, "processed", "vcontact_results.tsv"))

message("Database loading templates ready")
```

## 4. Host Prediction Analysis

### 4.1 Using ViroProfiler Pipeline

Our existing ViroProfiler pipeline can be used for phage-host prediction.

```{r viroprofiler-template}
#| eval: false

# Template for running ViroProfiler
# ViroProfiler path (update as needed)
viroprofiler_path <- "~/tools/ViroProfiler"

# Command template
viroprofiler_cmd <- sprintf(
  "cd %s && bash run_viroprofiler.sh -i %s -o %s -t phage_host",
  viroprofiler_path,
  file.path(path_upstream, "raw", "phage_contigs.fasta"),
  file.path(path_target, "processed", "viroprofiler_output")
)

# Execute (when ready)
# system(viroprofiler_cmd)
```

### 4.2 Using vConTACT3

vConTACT3 for viral cluster-based host prediction.

```{r vcontact-template}
#| eval: false

# Template for vConTACT3 analysis
# Input: viral protein files
# Output: cluster network with host predictions

vcontact_cmd <- sprintf(
  "vcontact3 --input-file %s --output-dir %s --threads 8",
  file.path(path_upstream, "processed", "viral_proteins.faa"),
  file.path(path_target, "processed", "vcontact3_output")
)

# Execute (when ready)
# system(vcontact_cmd)
```

## 5. Results Integration

### 5.1 Parse Prediction Results

```{r parse-results}
#| eval: false

# Template: Parse ViroProfiler/vConTACT3 results
# results_df <- read_tsv(file.path(path_target, "processed", "predictions.tsv"))

# Summary table template
# summary_table <- results_df %>%
#   group_by(target_pathogen, phage_id) %>%
#   summarise(
#     prediction_score = mean(score),
#     n_tools = n_distinct(tool),
#     .groups = "drop"
#   ) %>%
#   arrange(desc(prediction_score))

# knitr::kable(summary_table, caption = "Phage-Host Prediction Results")
```

### 5.2 Visualization

```{r visualize-predictions}
#| eval: false

# Template: Visualize prediction results
# ggplot(summary_table, aes(x = target_pathogen, y = prediction_score, fill = n_tools)) +
#   geom_boxplot() +
#   theme_minimal() +
#   labs(title = "Phage-Host Prediction Scores by Pathogen",
#        x = "Target Pathogen", y = "Prediction Score")
```

## 6. Candidate Phage Selection

### 6.1 Selection Criteria

Based on the Research Master Plan:
1. **Lytic lifestyle** (not temperate)
2. **Broad host range** (target multiple strains)
3. **Genomic stability** (few lysogenic genes)
4. **Safety profile** (no toxin genes, no AMR genes)

### 6.2 Top Candidates Table

*To be populated after running predictions*

 Phage ID | Target Pathogen | Prediction Score | Lifestyle | Host Range | Notes |
-----------|-----------------|------------------|------------|-------------|-------|
 TBD | C. perfringens | - | lytic | - | - |
 TBD | S. enterica | - | lytic | - | - |
 TBD | E. coli | - | lytic | - | - |
 TBD | Vibrio spp. | - | lytic | - | - |

## 7. NotebookLM Integration

### 7.1 Query NotebookLM for Validation

Use NotebookLM to validate predictions against literature.

```bash
nlm notebook query phage-synergy "What phages are effective against Clostridium perfringens in poultry?"
```

### 7.2 Add Results to NotebookLM

After generating prediction results, add them as a source to NotebookLM for cross-referencing with literature.

```bash
# Convert results to text/PDF and add to NotebookLM
nlm source add phage-synergy --text "$(cat predictions_summary.txt)" --title "Phage-Host Predictions pr0117"
```

## 8. Session Info

```{r session-info}
sessionInfo()
```

## 9. Next Steps

1. **Run ViroProfiler** on target phage genomes
2. **Run vConTACT3** for cluster-based predictions
3. **Integrate results** from multiple tools
4. **Validate predictions** using NotebookLM literature
5. **Proceed to Module 03** (`03-probiotic-target-identification.qmd`)

---
*Generated by qproj workflow for pr0117-PhageProbioticSynergy*

