---
title: "Probiotic Target Identification"
params:
  name: "03-probiotic-target-identification"
---

# Probiotic Target Identification

## Setup

```{r setup}
here::i_am(paste0(params$name, ".qmd"), uuid = "c3d4e5f6-a7b8-9012-cdef-1234567890")

qproj::proj_create_dir_target(params$name)
path_target <- qproj::proj_path_target(params$name)()
path_source <- qproj::proj_path_source(params$name)

# Read upstream data
path_01 <- qproj::proj_path_source("01-public-data-mining")
path_02 <- qproj::proj_path_source("02-phage-host-prediction")
```

## Load Packages

```{r packages}
library(tidyverse)
library(here)
library(qproj)
library(ggplot2)
```

## 1. Objective

Identify which probiotic bacteria benefit most from phage-mediated niche clearing:

- **Target probiotics**: *Lactobacillus*, *Bifidobacterium*, *Bacillus*, *Enterococcus*
- **Goal**: Understand colonization requirements, receptor usage, metabolic dependencies
- **Outcome**: Rank probiotics by expected synergy with phages

## 2. Probiotic Colonization Requirements

### 2.1 Key Factors for Gut Colonization

Based on NotebookLM literature review:

 Factor | Description | Relevance to Synergy |
--------|-------------|---------------------|
 Adhesion molecules | Bind to gut epithelial cells | Phage clearing competitors → more adhesion sites |
 Metabolic flexibility | Utilize diverse carbon sources | Survive after pathogen removal |
 Bile tolerance | Survive gut environment | Essential for delivery |
 Competitive exclusion | Outcompete pathogens | Enhanced by phage predation |
 Immune modulation | Regulate host immune response | Complementary to phage action |

### 2.2 Receptor Landscape Analysis

Analyze what receptors/pathways probiotics use for colonization.

```{r receptor-analysis-template}
#| eval: false

# Template: Analyze probiotic genome for adhesion factors
# Key gene families to search:
# - Mub (mucin-binding protein)
# - SpaCBA (sortase-anchored pilus)
# - LBA (lactobacillus adhesin)
# - S-layer proteins

# Example using eggnog-mapper results (if available)
# probiotic_genes <- read_tsv(file.path(path_02, "processed", "probiotic_annotation.tsv"))

# Filter for adhesion-related genes
# adhesion_genes <- probiotic_genes %>%
#   filter(grepl("adhesin|mucin|pilus|pilin|S-layer", description, ignore.case = TRUE))

# knitr::kable(head(adhesion_genes, 10), caption = "Adhesion-Related Genes in Probiotics")
```

## 3. Probiotic Fitness Knowledge Graph (Strategy 3 Enhancement)

### 3.1 Knowledge Graph Construction

Build a knowledge graph integrating multiple databases for objective probiotic scoring.

```{python knowledge-graph}
#| eval: false

# Knowledge graph using NetworkX (Python)
import networkx as nx
import pandas as pd
import json

# Initialize knowledge graph
G = nx.DiGraph()

# === Node types ===
# 1. Probiotic strains (from ProbiMed/BacMap)
probiotics = pd.DataFrame({
    'strain': ['Lactobacillus_acidophilus_NCFM', 'Bifidobacterium_lactis_BB12', 
               'Bacillus_subtilis_NRRL_B-30809', 'Enterococcus_faecium_NCIMB_10415'],
    'genome_size': [1.9, 2.2, 4.2, 2.8],  # Mbp
    'GC_content': [35.2, 58.1, 43.5, 37.8],
    'origin': ['human', 'animal', 'soil', 'animal']
})

# Add probiotic nodes
for _, row in probiotics.iterrows():
    G.add_node(row['strain'], type='probiotic', **row.to_dict())

# 2. Functional modules (from KEGG/BacMap)
functional_modules = [
    ('bacteriocin_synthesis', 'pathway', {'category': 'antimicrobial'}),
    ('adhesion_proteins', 'pathway', {'category': 'colonization'}),
    ('bile_salt_hydrolase', 'pathway', {'category': 'survival'}),
    ('short_chain_fatty_acid', 'pathway', {'category': 'metabolism'}),
    ('CRISPR_Cas', 'pathway', {'category': 'defense'})
]

G.add_nodes_from([(m[0], m[2]) for m in functional_modules])

# 3.Edgetypes: has_pathway, inhibits, enhances_synergy
# Example edges (would be populated from real DB queries)
edges = [
    ('Lactobacillus_acidophilus_NCFM', 'adhesion_proteins', {'weight': 0.9}),
    ('Lactobacillus_acidophilus_NCFM', 'bacteriocin_synthesis', {'weight': 0.8}),
    ('bacteriocin_synthesis', 'pathogen_suppression', {'weight': 0.85}),
    ('adhesion_proteins', 'gut_colonization', {'weight': 0.9})
]

G.add_edges_from(edges)

print(f"Knowledge graph created: {G.number_of_nodes()} nodes, {G.number_of_edges()} edges")
```

### 3.2 Feature Engineering for ML Scoring

Extract quantifiable features from genomes and knowledge graph.

```{python feature-engineering}
#| eval: false

import numpy as np
from sklearn.preprocessing import StandardScaler

# Feature matrix (example - would be computed from real data)
features = pd.DataFrame({
    'strain': probiotics['strain'],
    'genome_size_mbp': probiotics['genome_size'],
    'gc_content': probiotics['GC_content'],
    'bacteriocin_genes': [5, 3, 8, 6],  # Count from genome annotation
    'adhesion_genes': [12, 8, 4, 10],
    'bile_tolerance_score': [0.9, 0.85, 0.6, 0.8],  # From experiments
    'gut_colonization_score': [0.85, 0.8, 0.5, 0.75],  # From animal trials
    'cr CRISPR_spacers': [15, 22, 8, 18],  # From CRISPR array analysis
    'niche_overlap_pathogen': [0.7, 0.6, 0.4, 0.65]  # Computed from niche modeling
})

# Target variable: synergy score (to be predicted/validated)
# This would come from literature + experimental validation
features['synergy_score'] = [0.82, 0.78, 0.55, 0.75]  # Placeholder

print("Feature matrix shape:", features.shape)
print(features.head())
```

### 3.3 XGBoost Model for Probiotic Ranking

Train ML model to predict probiotic synergy potential with hyperparameter tuning.

```{python xgboost-model}
#| eval: false

import xgboost as xgb
from sklearn.model_selection import train_test_split, cross_val_score, GridSearchCV
from sklearn.metrics import mean_squared_error, r2_score, mean_absolute_error
import pandas as pd
import numpy as np

# Prepare data
X = features.drop(['strain', 'synergy_score'], axis=1)
y = features['synergy_score']

# Train-test split
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.3, random_state=42)

# === Base model ===
model = xgb.XGBRegressor(
    n_estimators=100,
    max_depth=3,
    learning_rate=0.1,
    random_state=42
)
model.fit(X_train, y_train)

# Predictions
y_pred = model.predict(X_test)
mse = mean_squared_error(y_test, y_pred)
r2 = r2_score(y_test, y_pred)
mae = mean_absolute_error(y_test, y_pred)

print(f"Base Model Performance:")
print(f"  MSE: {mse:.4f}")
print(f"  R² Score: {r2:.4f}")
print(f"  MAE: {mae:.4f}")

# === Cross-validation ===
print(f"\n=== Cross-Validation (5-fold) ===")
cv_scores = cross_val_score(model, X, y, cv=5, scoring='r2')
print(f"  R² scores: {cv_scores}")
print(f"  Mean R²: {cv_scores.mean():.4f} (+/- {cv_scores.std():.4f})")

# === Hyperparameter Tuning ===
print(f"\n=== Hyperparameter Tuning (GridSearch) ===")
param_grid = {
    'n_estimators': [50, 100, 200],
    'max_depth': [3, 5, 7],
    'learning_rate': [0.01, 0.1, 0.3],
    'subsample': [0.8, 1.0],
    'colsample_bytree': [0.8, 1.0]
}

grid_search = GridSearchCV(
    xgb.XGBRegressor(random_state=42),
    param_grid,
    cv=5,
    scoring='r2',
    n_jobs=-1,
    verbose=1
)
grid_search.fit(X_train, y_train)

print(f"\nBest parameters: {grid_search.best_params_}")
print(f"Best cross-validation R²: {grid_search.best_score_:.4f}")

# Use best model
best_model = grid_search.best_estimator_
y_pred_best = best_model.predict(X_test)
r2_best = r2_score(y_test, y_pred_best)
print(f"Test R² with tuned model: {r2_best:.4f}")

# === Feature Importance (from best model) ===
importance = pd.DataFrame({
    'feature': X.columns,
    'importance': best_model.feature_importances_
}).sort_values('importance', ascending=False)

print(f"\nFeature Importance (Tuned Model):")
print(importance)

# === Save model ===
import joblib
joblib.dump(best_model, 'probiotic_synergy_model.pkl')
print(f"\n✓ Model saved to probiotic_synergy_model.pkl")
```

### 3.4 Visualization with Cytoscape

Visualize knowledge graph and feature importance.

```{r cytoscape-viz}
#| eval: false

# R interface to Cytoscape (if available)
if (requireNamespace("RCy3", quietly = TRUE)) {
  library(RCy3)
  
  # Create Cytoscape network from Python-computed graph
  # (Integration would require exporting G to CSV/JSON)
  
  # Example: Bar plot of feature importance
  library(ggplot2)
  
  ggplot(importance, aes(x = reorder(feature, importance), y = importance)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    coord_flip() +
    theme_minimal() +
    labs(title = "Feature Importance for Probiotic Synergy Prediction",
         x = "Feature", y = "Importance (XGBoost)")
}
```

## 4. Niche Competition Modeling

### 4.1 Ecological Niche Overlap

Model the competitive relationship between pathogens and probiotics.

```{r niche-overlap-template}
#| eval: false

# Template: Calculate niche overlap index
# Based on:
# - Carbon source utilization profiles
# - Oxygen tolerance
# - pH optima
# - Temperature range
# - Receptor usage

# Example data structure
niche_data <- tibble(
  species = c("C. perfringens", "L. acidophilus", "S. enterica", "B. subtilis"),
  temp_opt (c(37, 37, 37, 30)),  # Celsius
  pH_opt (c(6.5, 6.0, 7.0, 7.5)),
  oxygen = c("anaerobic", "microaerophilic", "facultative", "aerobic"),
  carbon_sources = c(4, 6, 8, 10)  # Number of carbon sources utilized
)

# Calculate overlap score (simplified)
# Overlap = 1 / (1 + |temp_diff| + |pH_diff| + |carbon_diff|)
```

### 3.2 Phage Impact Simulation

Simulate how phage predation on pathogens opens niche space for probiotics.

```{r phage-impact-simulation}
#| eval: false

# Template: Simple competition model
# Initial abundances
initial_state <- tibble(
  species = c("Pathogen", "Probiotic", "Commensal1", "Commensal2"),
  abundance = c(100, 10, 50, 40),
  growth_rate = c(1.2, 0.8, 0.9, 0.7),
  susceptibility = c(1.0, 0.0, 0.3, 0.2)  # Phage susceptibility
)

# Simulation parameters
phage_mortality <- 0.8  # Pathogen mortality due to phage

# Simulate after phage treatment
post_phage <- initial_state %>%
  mutate(
    new_abundance = ifelse(susceptibility > 0,
      abundance * (1 - susceptibility * phage_mortality),
      abundance * (1 + growth_rate * 0.1))
  )

# Visualize
# ggplot(post_phage, aes(x = species, y = new_abundance, fill = species)) +
#   geom_bar(stat = "identity") +
#   theme_minimal() +
#   labs(title = "Abundance After Phage Treatment",
#        y = "Relative Abundance", x = "")
```

## 4. Probiotic Ranking

### 4.1 Ranking Criteria

From the Research Master Plan:

 Criterion | Weight | Description |
-----------|--------|-------------|
 Niche overlap with pathogen | 0.3 | Higher overlap = more benefit from phage |
 Lytic phage availability | 0.25 | Ensure effective phages exist |
 Commercial availability | 0.2 | Ready for application |
 Safety profile | 0.15 | GRAS status, no virulence |
 Mechanism clarity | 0.1 | Understanding of synergy |

### 4.2 Preliminary Ranking

*To be populated after analysis*

 Rank | Probiotic | Target Pathogen | Niche Overlap | Phage Available | Score |
------|------------|-----------------|---------------|------------------|-------|
 1 | TBD | TBD | - | - | - |
 2 | TBD | TBD | - | - | - |
 3 | TBD | TBD | - | - | - |

## 5. Genomic Analysis

### 5.1 Probiotic Genome Features

Analyze probiotic genomes for synergy-relevant features.

```{r genome-features-template}
#| eval: false

# Template: Analyze probiotic genomes
# Load probiotic genome annotations
# probiotic_gff <- rtracklayer::import.gff3(file.path(path_02, "raw", "probiotic_genomes.gff"))

# Extract features of interest
# - Bacteriocin genes (competitive advantage)
# - Carbohydrate utilization loci (niche exploitation)
# - Stress response genes (survival)
# - Bile salt hydrolase (gut adaptation)
```

### 5.2 Comparative Genomics

Compare probiotic strains for optimal selection.

```{r comparative-genomics}
#| eval: false

# Template: Compare multiple strains of same species
# Identify core and accessory genome
# Focus on strains with proven efficacy in animal trials
```

## 6. NotebookLM Validation

### 6.1 Query for Probiotic Evidence

Use NotebookLM to validate probiotic selection.

```bash
nlm notebook query phage-synergy "Which Lactobacillus strains show best colonization in poultry gut?"
```

### 6.2 Add Analysis Results

After generating ranking, add to NotebookLM for literature cross-validation.

```bash
nlm source add phage-synergy --text "$(cat probiotic_ranking.txt)" --title "Probiotic Ranking pr0117"
```

## 7. Session Info

```{r session-info}
sessionInfo()
```

## 8. Next Steps

1. **Complete genomic analysis** of top probiotic candidates
2. **Finalize ranking** based on multi-criteria assessment
3. **Cross-validate** with NotebookLM literature
4. **Proceed to Module 04** (`04-synergy-scoring.qmd`)

---
*Generated by qproj workflow for pr0117-PhageProbioticSynergy*

