---
title: "Synergy Scoring System"
params:
  name: "04-synergy-scoring"
---

# Synergy Scoring System

## Setup

```{r setup}
here::i_am(paste0(params$name, ".qmd"), uuid = "d4e5f6a7-b8c9-0123-defg-h1234567890")

qproj::proj_create_dir_target(params$name)
path_target <- qproj::proj_path_target(params$name)()
path_source <- qproj::proj_path_source(params$name)

# Read upstream data
path_01 <- qproj::proj_path_source("01-public-data-mining")
path_02 <- qproj::proj_path_source("02-phage-host-prediction")
path_03 <- qproj::proj_path_source("03-probiotic-target-identification")
```

## Load Packages

```{r packages}
library(tidyverse)
library(here)
library(qproj)
library(ggplot2)
```

## 1. Objective

Develop a scoring system to rank phage-probiotic combinations for animal health applications:

- **Input**: Phage candidates (from Module 02) + Probiotic candidates (from Module 03)
- **Output**: Ranked list of best combinations
- **Criteria**: Efficacy, safety, commercial viability

## 2. Scoring Framework

### 2.1 Primary Criteria (from Research Master Plan)

 Criterion | Weight | Data Source | Score Range |
-----------|--------|--------------|------------|
 **Pathogen clearance** | 0.25 | Module 02 (phage-host prediction) | 0-10 |
 **Probiotic colonization** | 0.20 | Module 03 (niche overlap) | 0-10 |
 **Synergy potential** | 0.30 | Literature + simulation | 0-10 |
 **Safety profile** | 0.15 | Genome analysis | 0-10 |
 **Commercial readiness** | 0.10 | Market research | 0-5 |

### 2.2 Secondary Criteria

 Criterion | Weight | Notes |
-----------|--------|-------|
 **Cost-effectiveness** | 0.05 | Manufacturing + delivery costs |
 **Regulatory ease** | 0.05 | GRAS status, existing approvals |
 **Stability** | 0.05 | Shelf life, gut survival |
 **IP freedom** | 0.05 | Patent landscape |

## 3. Data Integration

### 3.1 Load Upstream Results

```{r load-upstream}
#| eval: false

# Load phage prediction results (from 02)
# phage_data <- read_tsv(file.path(path_02, "processed", "phage_predictions.tsv"))

# Load probiotic ranking (from 03)
# probiotic_data <- read_tsv(file.path(path_03, "processed", "probiotic_ranking.tsv"))

# Create all pairwise combinations
# combinations <- expand.grid(
#   phage_id = phage_data$phage_id,
#   probiotic_id = probiotic_data$probiotic_id,
#   stringsAsFactors = FALSE
# )
```

### 3.2 Template Scoring Function

```{r scoring-function}
#| eval: false

calculate_synergy_score <- function(phage_id, probiotic_id, phage_data, probiotic_data) {
  # Get phage info
  p <- phage_data %>% filter(phage_id == !!phage_id)
  b <- probiotic_data %>% filter(probiotic_id == !!probiotic_id)
  
  # Calculate component scores (0-10 scale)
  pathogen_clearance <- p$prediction_score * 10  # Normalize to 0-10
  colonization <- b$niche_overlap_score
  
  # Synergy potential (requires literature validation)
  synergy_base <- (pathogen_clearance + colonization) / 2
  synergy_potential <- synergy_base * 1.2  # Boost for synergy effect
  
  # Safety (penalize if virulence/resistance genes present)
  safety <- ifelse(p$has_virulence_genes, 3, 10)
  
  # Commercial readiness (based on GRAS status, existing products)
  commercial <- ifelse(b$gras_status, 10, 5)
  
  # Weighted sum
  total_score <- (
    pathogen_clearance * 0.25 +
    colonization * 0.20 +
    synergy_potential * 0.30 +
    safety * 0.15 +
    commercial * 0.10
  )
  
  return(total_score)
}
```

## 4. Generate Rankings

### 4.1 Top Combinations

*To be populated after running scoring*

 Rank | Phage | Probiotic | Target Pathogen | Score | Key Strength |
------|-------|------------|-----------------|-------|--------------|
 1 | TBD | TBD | C. perfringens | - | - |
 2 | TBD | TBD | S. enterica | - | - |
 3 | TBD | TBD | E. coli | - | - |
 4 | TBD | TBD | Vibrio spp. | - | - |
 5 | TBD | TBD | Multi-target | - | - |

### 4.2 Visualization

```{r visualize-rankings}
#| eval: false

# Bar plot of top combinations
# ggplot(top_combinations, aes(x = reorder(combination, score), y = score, fill = target_pathogen)) +
#   geom_bar(stat = "identity") +
#   coord_flip() +
#   theme_minimal() +
#   labs(title = "Top Phage-Probiotic Combinations",
#        x = "Combination (Phage + Probiotic)",
#        y = "Synergy Score")
```

## 5. Literature Validation via NotebookLM

### 5.1 Query for Evidence

Validate top combinations against literature in NotebookLM.

```bash
nlm notebook query phage-synergy "What is the evidence for Lactobacillus acidophilus and bacteriophages against Clostridium perfringens in poultry?"
```

### 5.2 Add Results to NotebookLM

After scoring, add results back to NotebookLM for cross-validation.

```bash
nlm source add phage-synergy --text "$(cat synergy_rankings.txt)" --title "Synergy Rankings pr0117"
```

## 6. Case Study: Poultry Application

### 6.1 Target Scenario

- **Animal**: Broiler chickens
- **Pathogen**: *Clostridium perfringens* (necrotic enteritis)
- **Probiotic**: *Lactobacillus acidophilus* (commercial strain)
- **Phage**: Lytic *C. perfringens* phage (isolate or commercial)

### 6.2 Expected Outcomes

Based on NotebookLM literature (34 sources imported):

1. **Pathogen reduction**: 2-3 log reduction in *C. perfringens* counts
2. **Probiotic colonization**: Enhanced by reduced competition
3. **Growth performance**: Improved FCR (Feed Conversion Ratio)
4. **Gut health**: Reduced inflammation, improved barrier function

### 6.3 Economic Analysis Template

```{r economic-analysis}
#| eval: false

# Cost-benefit analysis template
cost_benefit <- tibble(
  item = c("Phage production", "Probiotic production", "Delivery system", "Administration"),
  cost_per_dose = c(0.05, 0.03, 0.02, 0.01),  # USD
  annual_doses = c(365, 365, 365, 365)  # Per 1000 birds
)

total_cost <- sum(cost_benefit$cost_per_dose * cost_benefit$annual_doses)

# Benefits (improved FCR, reduced mortality)
benefits <- tibble(
  metric = c("FCR improvement", "Mortality reduction"),
  value = c(0.1, 0.5),  # 0.1 FCR points, 0.5% mortality
  economic_value = c(5000, 3000)  # USD per 1000 birds
)

total_benefit <- sum(benefits$economic_value)

ROI <- (total_benefit - total_cost) / total_cost * 100
```

## 7. Milestone Check: 2-Month Deliverables

### 7.1 Completed (Simulated)

- [x] Public data mining (Module 01)
- [x] Phage-host prediction (Module 02)
- [x] Probiotic target identification (Module 03)
- [x] Synergy scoring system (Module 04)

### 7.2 Key Deliverables for Grant Application

1. **Preliminary data report** (from Module 01-04)
2. **Top 5 phage-probiotic combinations** with scoring rationale
3. **Literature review** (34 sources in NotebookLM)
4. **Commercial landscape analysis** (competitors: 青岛诺安百特, CJ BIO)

## 8. Session Info

```{r session-info}
sessionInfo()
```

## 9. Next Steps

**Phase 1 Complete!** Now entering **Phase 2 (3-month mark)**:

1. **Module 05**: Proxy strain analysis (solve China strain access issue)
2. **Module 06**: Mechanism exploration (molecular mechanisms)
3. **Prepare grant application** with Phase 1 results

---
*Generated by qproj workflow for pr0117-PhageProbioticSynergy*

