---
title: "Proxy Strain Analysis for China Market"
params:
  name: "05-proxy-strain-analysis"
---

# Proxy Strain Analysis for China Market

## Setup

```{r setup}
params <- list(name = "05-proxy-strain-analysis")
here::i_am(paste0(params$name, ".qmd"), uuid = "e5f6a7b8-c9d0-1234-efgh-i1234567890")

qproj::proj_create_dir_target(params$name)
path_target <- qproj::path_target(params$name)
path_source <- qproj::path_source(params$name)

# Read upstream data
path_01 <- qproj::path_source("01-public-data-mining")
path_02 <- qproj::path_source("02-phage-host-prediction")
path_04 <- qproj::path_source("04-synergy-scoring")
```

## Load Packages

```{r packages}
library(tidyverse)
library(here)
library(qproj)
library(ape)
library(ggplot2)
library(ggtree)
```

## 1. Problem Statement

### 1.1 The "Proxy Strain" Challenge

**核心问题**：无法在德国获取中国本地病原体菌株，但需要在德国进行湿实验验证。

**Notion 文档 Section 6.2**：
- 由于无法获取中国本地病原体菌株，策略是识别**可获得的国际参考菌株**
- 关键约束：代理菌株必须与中国本地菌株在基因组和表型上高度相似

### 1.2 Target Pathogens for China Market

| Pathogen | Disease | Target Animal | China Market Priority |
|----------|---------|---------------|----------------------|
| *Clostridium perfringens* | 坏死性肠炎 | 家禽 (poultry) | ⭐⭐⭐⭐⭐ |
| *Salmonella enterica* | 沙门氏菌病 | 家禽/牲畜 | ⭐⭐⭐⭐ |
| *Escherichia coli* (pathogenic) | 大肠杆菌病 | 家禽/水产 | ⭐⭐⭐⭐ |
| *Vibrio parahaemolyticus* | 急性肠胃炎 | 水产 (aquaculture) | ⭐⭐⭐ |
| *Streptococcus suis* | 猪链球菌病 | 生猪 (pig) | ⭐⭐⭐ |

## 2. Proxy Strain Selection Strategy

### 2.1 Selection Criteria

Based on Notion Research Master Plan:

1. **Genomic similarity** > 95% ANI (Average Nucleotide Identity)
2. **Phage susceptibility** (similar phage receptor profiles)
3. **Availability** (can obtain from DSMZ, ATCC, etc.)
4. **Safety** (moderate virulence for lab work)
5. **Publication track record** (well-characterized in literature)

### 2.2 Data Sources for Strain Comparison

```{r data-sources}
#| eval: false

# Template: Search NCBI for available strains
if (FALSE) {
  library(rentrez)
  
  # Search for C. perfringens strains in NCBI BioSample
  search_query <- 'Clostridium perfringens[Organism] AND "China"[Geolocation] AND ("strain" OR "isolate") AND ("genome" OR "complete genome")'
  
  strains_search <- rentrez::entrez_search(
    db = "biosample",
    term = search_query,
    retmax = 100
  )
  
  # Save for analysis
  write_lines(
    strains_search$ids,
    file.path(path_target, "raw", "c_perfringens_china_strains.txt")
  )
}
```

## 3. Strain Comparison Analysis

### 3.1 ANI Calculation Template

```{r ani-template}
#| eval: false

# Template for ANI calculation using pyani (Python) or fastANI
# Install: conda install -c bioconda pyani fastani

ani_script <- c(
  "#!/bin/bash",
  "# ANI calculation script",
  paste0("TARGET_DIR=", path_target, "/processed/ani_results"),
  "mkdir -p $TARGET_DIR",
  "",
  "# Run fastANI",
  "# fastANI -q query_genome.fasta -r reference_genome.fasta -o $TARGET_DIR/ani_output.txt",
  "",
  "# Batch ANI for multiple strains",
  "# for q in strains/*.fasta; do",
  "#   for r in reference/*.fasta; do",
  "#     fastANI -q $q -r $r -o $TARGET_DIR/$(basename $q)_vs_$(basename $r).txt",
  "#   done",
  "# done"
)

write_lines(
  ani_script,
  file.path(path_target, "scripts", "run_ani.sh")
)

# Make executable
Sys.chmod(file.path(path_target, "scripts", "run_ani.sh"), "755")
```

### 3.2 Phage Receptor Profile Comparison

```{r receptor-comparison}
#| eval: false

# Template: Compare phage receptor genes
# Key receptors to check:
# - LPS (Lipopolysaccharide) biosynthesis genes
# - T4P (Type IV Pilus) genes
# - Cell wall teichoic acid genes
# - Capsular polysaccharide genes

# Load genome annotations (from eggNOG-mapper or Prokka)
# query_annotation <- read_tsv(file.path(path_02, "processed", "query_annotation.tsv"))
# reference_annotation <- read_tsv(file.path(path_02, "processed", "reference_annotation.tsv"))

# Extract receptor-related genes
# receptor_genes <- c("lpx", "waa", "pil", "tat", "cap", "wzy")

# Compare presence/absence
# comparison_table <- bind_rows(
#   query_annotation %>% mutate(type = "query"),
#   reference_annotation %>% mutate(type = "reference")
# ) %>%
#   filter(grepl(paste(receptor_genes, collapse = "|"), gene_id, ignore.case = TRUE)) %>%
#   group_by(type, gene_id) %>%
#   summarise(count = n(), .groups = "drop")

# knitr::kable(comparison_table, caption = "Phage Receptor Gene Comparison")
```

## 4. Proxy Strain Candidates

### 4.1 *Clostridium perfringens* Proxies

*To be populated after analysis*

| Strain | Origin | ANI to China Isolate | Phage Susceptibility | Availability | Score |
|--------|--------|---------------------|----------------------|--------------|-------|
| ATCC 13124 | USA | - | Known | ATCC/DSMZ | - |
| DSM 756 | Germany | - | Moderate | DSMZ | - |
| TYPE B STRAIN | - | - | - | - | - |

### 4.2 *Salmonella enterica* Proxies

| Strain | Serovar | ANI to China Isolate | Phage Susceptibility | Availability | Score |
|--------|---------|---------------------|----------------------|--------------|-------|
| ATCC 14028 | Typhimurium | - | High | ATCC | - |
| DSM 554 | Enteritidis | - | Moderate | DSMZ | - |

### 4.3 *Vibrio* spp. Proxies (Aquaculture)

| Strain | Species | Origin | ANI | Host Range | Score |
|--------|---------|--------|-----|------------|-------|
| ATCC 17802 | V. parahaemolyticus | USA | - | Known | - |
| DSM 10027 | V. cholerae | Germany | - | Moderate | - |

## 5. Experimental Validation Plan

### 5.1 Proxy Validation Workflow

Based on Notion document Section 7 (The Bioinformatics-First Strategy):

**Step 1**: In silico validation
- ANI > 95% to China isolates
- Phage receptor profile match > 80%
- Metabolic capability similarity

**Step 2**: In vitro validation (Germany)
- Phage isolation using proxy strain
- Efficiency of plating (EOP) test
- Lytic spectrum confirmation

**Step 3**: Technology transfer to China
- Send validated phages to China collaborator
- Validate on local China isolates
- Compare efficacy (German proxy vs. China local)

### 5.2 Validation Metrics

```{r validation-metrics}
#| eval: false

validation_criteria <- tibble(
  metric = c("ANI similarity", "Receptor match", "Phage susceptibility", "Lytic spectrum", "Growth inhibition"),
  threshold = c("> 95%", "> 80%", "> 90%", "> 5 strains", "> 2 log reduction"),
  method = c("fastANI", "Gene presence/absence", "EOP assay", "Spot test", "In vitro co-culture")
)

knitr::kable(validation_criteria, caption = "Proxy Strain Validation Criteria")
```

## 6. Commercial Implications

### 6.1 China Market Access Strategy

From Notion document:
- **Competing companies**: 青岛诺安百特 (Qingdao Nuo'anbai), CJ BIO (Korea)
- **Strategy**: Use German lab for R&D, China partner for trials + commercialization
- **Timeline**: Proxy validation (3 months) → China validation (6 months) → Market entry (12 months)

### 6.2 IP Considerations

- Patent strategy must cover **both** proxy and local strain applications
- Phage cocktails should be broad enough to cover strain variations
- Probiotic combinations need to be effective across strain spectra

## 7. NotebookLM Validation

### 7.1 Query for Evidence

Use NotebookLM to validate proxy strain strategy:

```bash
nlm notebook query phage-synergy "What proxy strains are used for Clostridium perfringens research in Europe?"
```

### 7.2 Add Results to NotebookLM

After completing proxy analysis, add to NotebookLM:

```bash
nlm source add phage-synergy --text "$(cat proxy_strain_report.txt)" --title "Proxy Strain Analysis pr0117"
```

## 8. Session Info

```{r session-info}
sessionInfo()
```

## 9. Next Steps

1. **Complete ANI analysis** for top candidate strains
2. **Validate phage receptor profiles** against literature
3. **Order proxy strains** from DSMZ/ATCC
4. **Design validation experiments** (Germany lab)
5. **Proceed to Module 06** (`06-mechanism-exploration.qmd`)

---
*Generated by qproj workflow for pr0117-PhageProbioticSynergy*

