---
title: "Mechanism Exploration"
params:
  name: "06-mechanism-exploration"
---

# Mechanism Exploration

## Setup

```{r setup}
params <- list(name = "06-mechanism-exploration")
here::i_am(paste0(params$name, ".qmd"), uuid = "f6a7b8c9-d0e1-2345-ghij-k1234567890")

qproj::proj_create_dir_target(params$name)
path_target <- qproj::path_target(params$name)
path_source <- qproj::path_source(params$name)

# Read upstream data
path_02 <- qproj::path_source("02-phage-host-prediction")
path_03 <- qproj::path_source("03-probiotic-target-identification")
path_05 <- qproj::path_source("05-proxy-strain-analysis")
```

## Load Packages

```{r packages}
library(tidyverse)
library(here)
library(qproj)
library(ggplot2)
library(ape)
library(phyloseq)  # For microbiome analysis
# library(DESeq2)   # For differential abundance
# library(vegan)     # For ordination
```

## 1. Objective

Explore the molecular mechanisms by which phages enhance probiotic colonization:

- **Primary mechanism**: Niche clearing (pathogen removal)
- **Secondary mechanisms**: 
  - Receptor landscape changes
  - Metabolic shifts in gut environment
  - Immune modulation (host response)
  - Horizontal gene transfer (HGT) events

## 2. Mechanism Hypotheses

### 2.1 From NotebookLM Literature Review (34 sources)

Based on the 34 sources imported to NotebookLM:

 Mechanism | Evidence Strength | Key Papers |
-----------|-------------------|-------------|
 **Direct niche clearing** | Strong (15+ papers) | Effects of Phage Cocktail... (Broiler Chickens) |
 **Metabolic cross-feeding** | Moderate (8 papers) | Probiophage: Novel Candidate... |
 **Immune modulation** | Emerging (5 papers) | Engineered probiotics for IBD... |
 **Receptor remodeling** | Theoretical (3 papers) | Global research trends in bacteriophage... |
 **Biofilm disruption** | Moderate (6 papers) | Phage-Antibiotic Synergy Review... |

### 2.2 Testable Predictions

1. **Phage treatment reduces pathogen abundance** → Less competition for probiotics
2. **Probiotic metabolic activity increases** → Better colonization and persistence
3. **Gut inflammation decreases** → More favorable environment for probiotics
4. **Microbial diversity increases** → More stable ecosystem

## 3. Receptor Landscape Analysis

### 3.1 Pre- and Post-Phage Treatment

Compare receptor profiles before and after phage intervention.

```{r receptor-landscape}
#| eval: false

# Template: Load genomic data for pathogens and probiotics
# pathogen_genomes <- readDNAStringSet(file.path(path_02, "processed", "pathogen_genomes.fasta"))
# probiotic_genomes <- readDNAStringSet(file.path(path_03, "processed", "probiotic_genomes.fasta"))

# Key receptor genes to track:
receptor_genes <- tibble(
  receptor = c("LPS", "T4P", "Teichoic_acid", "Capsule", "S-layer", "Pilus"),
  gene_family = c("lpx", "pil", "tag", "cap", "slp", "pil"),
  target = c("Gram-negative", "Both", "Gram-positive", "Both", "Gram-positive", "Both")
)

# Simulate pre/post treatment (template)
# pre_treatment <- tibble(
#   sample = rep(c("pathogen", "probiotic"), each = 10),
#   receptor_expr = rnorm(20, mean = 50, sd = 10),
#   treatment = "pre"
# )
# 
# post_treatment <- tibble(
#   sample = rep(c("pathogen", "probiotic"), each = 10),
#   receptor_expr = c(rnorm(10, mean = 10, sd = 5), rnorm(10, mean = 70, sd = 10)),
#   treatment = "post"
# )
# 
# combined <- bind_rows(pre_treatment, post_treatment)
# 
# ggplot(combined, aes(x = treatment, y = receptor_expr, fill = sample)) +
#   geom_boxplot() +
#   facet_wrap(~receptor) +
#   theme_minimal() +
#   labs(title = "Receptor Expression Pre/Post Phage Treatment")
```

## 4. Metabolic Pathway Analysis

### 4.1 Metabolic Model Integration

Explore how phage-mediated pathogen removal affects metabolic environment.

```{r metabolic-analysis}
#| eval: false

# Template: Load KEGG pathways / MetaCyc data
# pathogen_metabolism <- readRDS(file.path(path_02, "processed", "pathogen_pathways.rds"))
# probiotic_metabolism <- readRDS(file.path(path_03, "processed", "probiotic_pathways.rds"))

# Key metabolic shifts to monitor:
metabolic_shifts <- tibble(
  pathway = c("Carbon metabolism", "Amino acid synthesis", "Bile acid transformation", 
               "Short-chain fatty acid production", "Vitamin synthesis"),
  pathogen_role = c("Consumer", "Consumer", "Converter", "Inhibitor", "Consumer"),
  probiotic_role = c("Consumer", "Producer", "Converter", "Producer", "Producer"),
  post_phage_change = c("↑ Probiotic", "↑ Probiotic", "↑ Probiotic", "↑ Probiotic", "↑ Probiotic")
)

knitr::kable(metabolic_shifts, caption = "Expected Metabolic Shifts After Phage Treatment")
```

### 4.2 Carbon Source Utilization

Compare carbon source utilization profiles.

```{r carbon-utilization}
#| eval: false

# Template: Using Biolog plate data or genomic prediction
# carbon_sources <- c("glucose", "fructose", "lactose", "cellulose", "xylan")

# Simulate utilization scores
# utilization_matrix <- tibble(
#   organism = rep(c("Pathogen", "Probiotic", "Commensal"), each = 5),
#   carbon_source = rep(carbon_sources, 3),
#   utilization_score = c(rnorm(5, 60, 10), rnorm(5, 80, 15), rnorm(5, 70, 12))
# )
# 
# ggplot(utilization_matrix, aes(x = carbon_source, y = utilization_score, fill = organism)) +
#   geom_bar(stat = "identity", position = "dodge") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#   labs(title = "Carbon Source Utilization Profiles")
```

## 5. Immune Modulation Analysis

### 5.1 Host Immune Response

Explore how the phage-probiotic combination modulates host immunity.

```{r immune-response}
#| eval: false

# Template: Cytokine profiling (from literature or experiments)
cytokines <- tibble(
  cytokine = c("IL-10", "TGF-β", "IL-6", "TNF-α", "IL-1β"),
  type = c("Anti-inflammatory", "Anti-inflammatory", "Pro-inflammatory", "Pro-inflammatory", "Pro-inflammatory"),
  pre_treatment = c(20, 15, 80, 90, 70),
  post_phage = c(25, 18, 60, 50, 40),  # Improved (↓ pro-inflammatory)
  post_combination = c(35, 25, 30, 25, 20)  # Best (↑ anti-inflammatory)
)

# Reshape for plotting
cytokine_long <- cytokines %>%
  pivot_longer(cols = c(pre_treatment, post_phage, post_combination),
               names_to = "treatment", values_to = "level")

# ggplot(cytokine_long, aes(x = cytokine, y = level, fill = treatment)) +
#   geom_bar(stat = "identity", position = "dodge") +
#   theme_minimal() +
#   labs(title = "Cytokine Levels Under Different Treatments",
#        y = "Relative Expression Level")
```

## 6. Horizontal Gene Transfer (HGT) Analysis

### 6.1 Phage-Mediated HGT Events

Track potential gene transfer between phages, pathogens, and commensals.

```{r hgt-analysis}
#| eval: false

# Template: Use PHASTER or similar tools to detect prophages
# prophage_regions <- read_tsv(file.path(path_02, "processed", "prophage_predictions.tsv"))

# Look for:
# - Antimicrobial resistance (AMR) genes
# - Virulence factors
# - Metabolic genes
# - Probiotic-associated genes

# hgt_events <- tibble(
#   donor = c("Pathogen", "Pathogen", "Probiotic"),
#   recipient = c("Commensal", "Probiotic", "Pathogen"),
#   gene_type = c("AMR", "Virulence", "Metabolic"),
#   frequency = c(0.05, 0.02, 0.01)  # Events per generation
# )
# 
# knitr::kable(hgt_events, caption = "Potential HGT Events")
```

## 7. Integrated Model

### 7.1 Conceptual Diagram

Create a systems-level model of phage-probiotic synergy.

**Components**:
1. **Phage predation** → Pathogen reduction
2. **Niche clearance** → Resource liberation
3. **Probiotic colonization** → Fills empty niches
4. **Metabolic synergy** → Enhanced gut health
5. **Immune modulation** → Reduced inflammation

### 7.2 Mathematical Model Template

```{r mathematical-model}
#| eval: false

# Template: Simple Lotka-Volterra competition model with phage predation
# Parameters:
# - r_p: pathogen growth rate
# - r_b: probiotic growth rate
# - a_pb: competition coefficient (pathogen on probiotic)
# - a_bp: competition coefficient (probiotic on pathogen)
# - K: carrying capacity
# - d_p: phage-induced pathogen death rate
# - m: phage maintenance cost

# Simulate over time
time <- seq(0, 100, by = 0.1)
N_p <- numeric(length(time))  # Pathogen abundance
N_b <- numeric(length(time))  # Probiotic abundance
N_p[1] <- 100; N_b[1] <- 10  # Initial conditions

for (t in 1:(length(time)-1)) {
  dN_p <- N_p[t] * (0.8 * (1 - (N_p[t] + 0.5*N_b[t])/100) - 0.3)  # With phage mortality
  dN_b <- N_b[t] * (0.6 * (1 - (N_b[t] + 0.3*N_p[t])/100))
  
  N_p[t+1] <- max(0, N_p[t] + dN_p)
  N_b[t+1] <- max(0, N_b[t] + dN_b)
}

# plot(time, N_p, type = "l", col = "red", ylim = c(0, 120),
#      xlab = "Time", ylab = "Abundance",
#      main = "Pathogen-Probiotic Dynamics with Phage Predation")
# lines(time, N_b, col = "blue")
# legend("topright", legend = c("Pathogen", "Probiotic"),
#        col = c("red", "blue"), lty = 1)
```

## 8. NotebookLM Validation

### 8.1 Query for Mechanism Evidence

Use NotebookLM to validate mechanism hypotheses:

```bash
nlm notebook query phage-synergy "What are the molecular mechanisms of phage-probiotic synergy in animal gut?"
```

### 8.2 Add Results to NotebookLM

After mechanism analysis, add findings to NotebookLM:

```bash
nlm source add phage-synergy --text "$(cat mechanism_summary.txt)" --title "Mechanism Exploration pr0117"
```

## 9. Session Info

```{r session-info}
sessionInfo()
```

## 10. Next Steps

1. **Run simulations** (Section 7.2) with real parameter estimates
2. **Validate with literature** from NotebookLM (34 sources)
3. **Design experiments** to test key mechanisms
4. **Prepare for Phase 3** (6-month mark: experimental design, manuscript)
5. **Proceed to Module 07** (`07-experimental-design.qmd`)

---
*Generated by qproj workflow for pr0117-PhageProbioticSynergy*

