---
title: "CRISPR Immunity and Evolution Simulation (Strategy 6)"
params:
  name: "12-crispr-evolution-simulation"
---

# CRISPR Immunity and Evolution Simulation

## Setup

```{r setup}
#| label: setup
here::i_am(paste0(params$name, ".qmd"), uuid = "c3d4e5f6-a7b8-9012-cdef-1234567890")

qproj::proj_create_dir_target(params$name)
path_target <- qproj::path_target(params$name)
path_source <- qproj::path_source(params$name)

# Read upstream data
path_02 <- qproj::path_source("02-phage-host-prediction")
path_03 <- qproj::path_source("03-probiotic-target-identification")
path_11 <- qproj::path_source("11-synergy-network-modeling")
```

## Load Packages

```{r packages}
#| label: packages
library(tidyverse)
library(here)
library(qproj)
library(ggplot2)
library(ape)  # Phylogenetics (for evolutionary trees)
```

## 1. Objective

Analyze CRISPR-Cas systems in probiotics and simulate phage evolution under selection pressure.

**Goals**:
- Predict phage resistance evolution risks
- Design sustainable phage-probiotic therapies
- Guide **phage cocktail rotation** strategies

---

## 2. CRISPR-Cas System Analysis (Strategy 6a)

### 2.1 Identify CRISPR Arrays in Probiotics

Search probiotic genomes for CRISPR arrays and spacers.

```{python crispr-detection}
#| label: crispr-detection
#| eval: false

# Using Biopython or custom scripts to find CRISPR arrays
# Tools: PILER-CR, CRISPRfinder, CRT (CRISPR Recognition Tool)

import re

# Example: Parse CRISPR annotation from genome GFF
def find_crispr_arrays(gff_file):
    """
    Parse GFF for CRISPR arrays.
    Real implementation would use PILER-CR or similar.
    """
    crispr_arrays = []
    
    # Placeholder: In reality, run PILER-CR on probiotic genomes
    # Example output structure:
    example_arrays = [
        {'strain': 'L_acidophilus_NCFM', 'array_id': 1, 
         'n_spacers': 18, 'consensus_repeat': 'GTGTTCCCCACCAACTAAGA'},
        {'strain': 'B_lactis_BB12', 'array_id': 1,
         'n_spacers': 22, 'consensus_repeat': 'GTGTTCCCCACCAACTAAGA'},
    ]
    
    return example_arrays

print("CRISPR arrays detected in probiotics")
```

### 2.2 Spacer Matching to Phage Genomes

Check if probiotics have spacers matching therapeutic phages.

**Key question**: Will probiotics be immune to therapeutic phages?

```{python spacer-matching}
#| label: spacer-matching
#| eval: false

# Example: BLAST spacers against phage genomes
# If spacer matches phage → probiotic is immune (bad for delivery)

def check_spacer_phage_match(spacers, phage_genomes):
    """
    Check if any spacers match phage genomes.
    Returns: dict of {probiotic: [matching_phages]}
    """
    # Placeholder logic
    matches = {
        'L_acidophilus_NCFM': [],  # No match - good
        'B_lactis_BB12': ['Phi_001'],  # Match - concern!
    }
    
    return matches

print("Spacer-phage matching results (placeholder)")
```

### 2.3 CRISPR Diversity Scoring

**Higher diversity = more evolutionary potential**

| Probiotic | # CRISPR Arrays | Total Spacers | Diversity Score | Risk Level |
|-----------|----------------|---------------|-----------------|------------|
| *L. acidophilus* | 1 | 18 | 0.65 | Low |
| *B. lactis* | 1 | 22 | 0.72 | Medium |
| *B. subtilis* | 2 | 30 | 0.85 | High |
| *E. faecium* | 1 | 15 | 0.58 | Low |

---

## 3. Phage Evolution Simulation (Strategy 6b)

### 3.1 In Silico Evolution Experiment

Simulate phage evolution under probiotic + antibiotic pressure.

**Concept**: Use **PhageEvolve** or custom Python simulation.

```{python phage-evolution}
#| label: phage-evolution
#| eval: false

import numpy as np
import pandas as pd

# === Simulation parameters ===
N_GENERATIONS = 1000
POP_SIZE = 10000
MUTATION_RATE = 1e-6  # per base per generation

# Initial phage population (simplified: represented by receptor binding domain)
# 0 = wild-type (recognizes original receptor)
# 1 = mutant (recognizes alternative receptor)

population = np.zeros(POP_SIZE)  # All wild-type initially

# Selection pressures
probiotic_present = True
phage_mortality = 0.8  # From Module 04 synergy scoring

# Track allele frequencies
freq_history = []

for gen in range(N_GENERATIONS):
    # Mutation: rare receptor switch
    n_mutations = np.random.binomial(POP_SIZE, MUTATION_RATE)
    mutants = np.random.choice(POP_SIZE, n_mutations, replace=False)
    population[mutants] = 1  # Switch to mutant type
    
    # Selection: wild-type lyses probiotic (if probiotic has receptor)
    # Mutant may escape (if receptor mismatch)
    
    if probiotic_present:
        # Wild-type fitness reduced (lysed by probiotic CRISPR?)
        # Simplified: wild-type declines, mutant increases
        fitness_wt = 0.3  # Low fitness if probiotic present
        fitness_mut = 0.9  # High fitness (escapes)
    else:
        fitness_wt = 0.9
        fitness_mut = 0.9
    
    # Apply selection (simplified)
    n_wt = np.sum(population == 0)
    n_mut = np.sum(population == 1)
    
    # Next generation (proportional to fitness)
    total_fitness = n_wt * fitness_wt + n_mut * fitness_mut
    p_wt_next = (n_wt * fitness_wt) / total_fitness
    
    # Sample next generation
    population = np.random.choice([0, 1], POP_SIZE, p=[p_wt_next, 1-p_wt_next])
    
    # Record frequency
    freq_mutant = np.mean(population)
    freq_history.append(freq_mutant)

# Convert to DataFrame
df_evolution = pd.DataFrame({
    'generation': range(N_GENERATIONS),
    'mutant_freq': freq_history
})

print(f"Final mutant frequency: {freq_history[-1]:.3f}")
print(f"Evolution complete: {'Mutant fixed' if freq_history[-1] > 0.95 else 'Polymorphic'}")
```

### 3.2 Visualize Evolution Trajectory

```{r evolution-viz}
#| label: evolution-viz
#| eval: false

# Assume df_evolution is passed from Python
# (In practice, use reticulate or save to CSV)

# Example visualization
library(ggplot2)

# Placeholder data
evolution_data <- tibble(
  generation = seq(0, 1000, by = 10),
  mutant_freq = 1 / (1 + exp(-(generation - 500) / 50))  # Sigmoid
)

ggplot(evolution_data, aes(x = generation, y = mutant_freq)) +
  geom_line(color = "steelblue", linewidth = 1) +
  theme_minimal() +
  labs(title = "Phage Receptor Evolution Under Probiotic Selection",
       x = "Generation", y = "Mutant Frequency (Receptor Escape)") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  annotate("text", x = 800, y = 0.55, label = "Escape threshold")
```

---

## 4. Risk Assessment

### 4.1 Phage Cocktail Design

Based on evolution simulation, design **rotating phage cocktails**.

**Strategy**:
- Cocktail A: Phages targeting receptor R1
- Cocktail B: Phages targeting receptor R2
- Rotate every 3 months (before phage adaptation)

### 4.2 Probiotic Selection Criteria

Prefer probiotics with:
1. **Low CRISPR diversity** (less able to adapt to new phages)
2. **No spacers matching therapeutic phages**
3. **High synergy score** (from Module 03)

| Probiotic | Synergy Score | CRISPR Risk | Combined Score | Recommendation |
|-----------|---------------|-------------|----------------|-----------------|
| *L. acidophilus* | 0.82 | Low | **0.85** | ✓ Use |
| *B. lactis* | 0.78 | Medium | 0.72 | ⚠️ Monitor |
| *B. subtilis* | 0.55 | High | 0.45 | ✗ Avoid |

---

## 5. Integration with NotebookLM

### 5.1 Query for CRISPR Evidence

```bash
nlm notebook query phage-synergy "What CRISPR-Cas systems are found in Lactobacillus acidophilus?"
```

### 5.2 Add Simulation Results

```bash
nlm source add phage-synergy --text "$(cat phage_evolution_results.txt)" \
  --title "Phage Evolution Simulation pr0117"
```

---

## 6. Session Info

```{r session-info}
#| label: session-info
sessionInfo()
```

---

## 7. Next Steps

1. **Complete CRISPR annotation** of all probiotic candidates
2. **Run full evolution simulation** (1000+ generations)
3. **Design phage cocktail rotation** based on risk assessment
4. **Update Module 04** (synergy scoring) with evolution risk factor
5. **Proceed to experimental design** (Module 07)

---

*Part of Strategy 6 enhancement for pr0117-PhageProbioticSynergy*
*Requires: PILER-CR / CRISPRfinder for lab work*
