---
title: "Aquaculture Pathogen Analysis (Extension)"
params:
  name: "13-aquaculture-pathogen-analysis"
---

# Aquaculture Pathogen Analysis

## Setup

```{r setup}
#| label: setup
here::i_am(paste0(params$name, ".qmd"), uuid = "c3d4e5f6-a7b8-9012-cdef-1234567890")

qproj::proj_create_dir_target(params$name)
path_target <- here::here("analyses", params$name, "target")
path_source <- here::here("analyses", params$name, "source")

# Read upstream data
path_02 <- here::here("analyses", "02-phage-host-prediction", "source")
path_03 <- here::here("analyses", "03-probiotic-target-identification", "source")
path_11 <- here::here("analyses", "11-synergy-network-modeling", "source")
```

## Load Packages

```{r packages}
#| label: packages
library(tidyverse)
library(here)
library(qproj)
library(ggplot2)
library(ggrepel)  # For labeling points
```

## 1. Objective

Focus on aquaculture-specific pathogens and optimize phage-probiotic synergy for aquatic farming.

**Aquaculture context**:
- **Target species**: Fish (tilapia, catfish, carp), shrimp, shellfish
- **Key pathogens**: *Aeromonas hydrophila*, *Vibrio spp.*, *Edwardsiella tarda*
- **Environmental factors**: Water temperature, salinity, pH, oxygen
- **Probiotic candidates**: *Bacillus spp.*, *Lactobacillus spp.*, *Enterococcus spp.*

---

## 2. Aquaculture Pathogen Database

### 2.1 Major Aquaculture Pathogens

| Pathogen | Disease | Host Range | Mortality | Phage Available |
|----------|---------|------------|-----------|------------------|
| *Aeromonas hydrophila* | Motile Aeromonad Septicemia | Freshwater fish | 30-70% | ✓ (Multiple) |
| *Vibrio harveyi* | Vibriosis | Marine fish/shrimp | 50-90% | ✓ (Some) |
| *Vibrio anguillarum* | Vibriosis | Marine fish | 40-80% | ✓ (Limited) |
| *Edwardsiella tarda* | Edwardsiellosis | Freshwater fish | 20-60% | ✗ (Rare) |
| *Streptococcus iniae* | Streptococcosis | Various fish | 30-50% | ✗ (None) |
| *Yersinia ruckeri* | Enteric Redmouth Disease | Trout | 10-40% | ✓ (Some) |

### 2.2 Pathogen Genomics

**Genome features relevant to phage therapy**:
- **Prophage load**: High prophage count → lysogenic phages (avoid)
- **CRISPR arrays**: Number of spacers → phage resistance risk
- **Capsule/ biofilm**: Phage accessibility issues
- **Antibiotic resistance**: Co-selection with phage resistance

```{python pathogen-genomics}
#| label: pathogen-genomics
#| eval: false

import pandas as pd

# Aquaculture pathogen database (example)
pathogen_db = pd.DataFrame({
    'species': ['Aeromonas_hydrophila', 'Vibrio_harveyi', 'Vibrio_anguillarum', 
                'Edwardsiella_tarda', 'Streptococcus_iniae'],
    'genome_size_mbp': [3.5, 5.8, 4.2, 3.8, 2.1],
    'gc_content': [61.5, 45.2, 44.8, 53.2, 36.5],
    'prophage_count': [3, 5, 2, 1, 0],  # Estimated
    'crispr_arrays': [2, 3, 2, 1, 0],
    'crispr_spacers': [28, 35, 22, 15, 0],
    'biofilm_forming': [True, True, True, False, True],
    'antibiotic_resistance': ['multi-drug', 'beta-lactam', 'multi-drug', 'quinolone', 'macrolide'],
    'aquaculture_host': ['freshwater', 'marine', 'marine', 'freshwater', 'various']
})

print("Aquaculture Pathogen Database:")
print(pathogen_db)

# Risk score: higher = more challenging for phage therapy
pathogen_db['phage_risk_score'] = (
    pathogen_db['prophage_count'] * 0.3 + 
    pathogen_db['crispr_spacers'] * 0.05 + 
    pathogen_db['biofilm_forming'].astype(int) * 0.5
)

pathogen_db_sorted = pathogen_db.sort_values('phage_risk_score', ascending=False)
print("\nPathogen Risk Ranking (higher = harder to treat with phages):")
print(pathogen_db_sorted[['species', 'phage_risk_score', 'aquaculture_host']])
```

---

## 3. Environmental Factors

### 3.1 Water Quality Parameters

**Critical factors affecting phage-probiotic synergy in aquaculture**:

| Factor | Optimal Range | Effect on Phage | Effect on Probiotic |
|--------|----------------|-------------------|-------------------|
| Temperature | 25-30°C | Affects phage replication rate | Affects probiotic growth |
| Salinity | 0-35 ppt | Marine phages vs freshwater | Probiotic salt tolerance |
| pH | 6.5-8.5 | Phage stability | Probiotic survival |
| Dissolved O₂ | >5 mg/L | Minimal | Essential for aerobes |
| Organic matter | <20 mg/L | Can inactivate phages | Nutrient source |

### 3.2 Simulate Environmental Effects

```{python env-simulation}
#| label: env-simulation
#| eval: false

import numpy as np
import pandas as pd

# Simulate phage stability under different conditions
def phage_stability_score(temp, salinity, pH):
    """
    Simplified model: phage stability index (0-1)
    Higher = more stable
    """
    # Temperature effect (optimal around 25-30°C)
    temp_score = 1.0 - abs(temp - 27.5) / 20.0
    temp_score = max(0, min(1, temp_score))
    
    # Salinity effect (marine phages more tolerant)
    salinity_score = 1.0 - salinity / 50.0
    salinity_score = max(0, min(1, salinity_score))
    
    # pH effect (optimal 7.0-8.0)
    ph_score = 1.0 - abs(pH - 7.5) / 5.0
    ph_score = max(0, min(1, ph_score))
    
    # Combined score (simplified)
    return (temp_score + salinity_score + ph_score) / 3

# Test different aquaculture scenarios
scenarios = [
    {'name': 'Freshwater pond', 'temp': 28, 'salinity': 0, 'pH': 7.2},
    {'name': 'Brackish water', 'temp': 26, 'salinity': 15, 'pH': 7.8},
    {'name': 'Marine cage', 'temp': 25, 'salinity': 35, 'pH': 8.1},
    {'name': 'Cold freshwater', 'temp': 15, 'salinity': 0, 'pH': 6.8}
]

results = []
for s in scenarios:
    score = phage_stability_score(s['temp'], s['salinity'], s['pH'])
    results.append({
        'scenario': s['name'],
        'temperature': s['temp'],
        'salinity': s['salinity'],
        'pH': s['pH'],
        'phage_stability': score
    })

df_scenarios = pd.DataFrame(results)
print("Phage Stability under Different Aquaculture Conditions:")
print(df_scenarios)
```

---

## 4. Probiotic Candidates for Aquaculture

### 4.1 Top Probiotic Species

Based on literature and Module 03 analysis, adapted for aquaculture:

| Probiotic | Target Pathogen | Aquaculture Use | Mechanism | Evidence |
|-----------|-----------------|------------------|------------|----------|
| *Bacillus subtilis* | *A. hydrophila* | ✓ Widely used | Competitive exclusion, bacteriocins | Strong |
| *Bacillus licheniformis* | *V. harveyi* | ✓ Common | Spore-former, stable in water | Moderate |
| *Lactobacillus plantarum* | *E. tarda* | ✓ Freshwater | Acidification, adhesion | Moderate |
| *Enterococcus faecium* | *S. iniae* | ✓ Some use | Competitive exclusion | Limited |
| *Pediococcus acidilactici* | Various | ✓ Emerging | Lactic acid, bacteriocins | Growing |

### 4.2 Aquaculture-Specific Scoring

**Additional criteria for aquaculture**:
- **Spore-forming ability** (for *Bacillus*): Survival in water
- **Salt tolerance**: For marine/brackish systems
- **Temperature tolerance**: 15-35°C range
- **Water column stability**: Not quickly sedimenting

---

## 5. Synergy Optimization for Aquaculture

### 5.1 Phage-Probiotic Pairing Strategy

**Strategy for different aquaculture systems**:

1. **Freshwater ponds** (tilapia, catfish):
   - Pathogen: *A. hydrophila*, *E. tarda*
   - Probiotic: *B. subtilis* (spore-former, stable)
   - Phage: *A. hydrophila*-specific (lytic, broad host range within species)

2. **Marine cages** (sea bass, sea bream):
   - Pathogen: *V. harveyi*, *V. anguillarum*
   - Probiotic: *B. licheniformis* (salt-tolerant)
   - Phage: *Vibrio*-specific (consider phage cocktails)

3. **Shrimp ponds**:
   - Pathogen: *V. harveyi*, *V. parahaemolyticus*
   - Probiotic: *B. subtilis* + *L. plantarum* (combination)
   - Phage: *Vibrio* phage cocktail (3-4 phages)

### 5.2 Dosing and Delivery

**Key considerations**:
- **Water volume**: Dose per m³ (phage: 10³-10⁶ PFU/L, probiotic: 10⁶-10⁸ CFU/L)
- **Delivery method**: Direct addition vs encapsulated (slow release)
- **Timing**: Probiotic first (3-5 days), then phage (establish niche)
- **Monitoring**: Weekly water sampling + qPCR for pathogen load

---

## 6. Risk Assessment for Aquaculture

### 6.1 Phage Resistance Evolution (Aquaculture Context)

**Higher risk factors**:
- High pathogen density in ponds (10⁶-10⁸ CFU/mL)
- Continuous phage exposure (vs pulsed)
- Sub-lethal phage doses (incomplete lysis)

**Mitigation strategies**:
- **Phage cocktails** (3-5 phages targeting different receptors)
- **Rotating phages** (every 2-3 weeks)
- **Combine with probiotics** (reduces selection pressure)
- **Monitor CRISPR spacers** (early warning system)

### 6.2 Regulatory and Safety

- **Phage approval**: USDA/FDA regulations (US), EFSA (EU)
- **Probiotic GRAS status**: Ensure aquaculture-approved strains
- **Environmental release**: Risk of phage dispersal to wild populations
- **Antibiotic interaction**: Avoid during antibiotic treatment

---

## 7. Integration with NotebookLM

### 7.1 Query for Aquaculture Evidence

```bash
nlm notebook query phage-synergy "What are the most effective probiotic strains for Vibrio control in shrimp aquaculture?"
```

### 7.2 Add Aquaculture Data

```bash
nlm source add phage-synergy --text "$(cat aquaculture_pathogen_db.csv)" \
  --title "Aquaculture Pathogen Database pr0117"
```

---

## 8. Session Info

```{r session-info}
#| label: session-info
sessionInfo()
```

---

## 9. Next Steps

1. **Validate pathogen database** with NotebookLM literature search
2. **Run Module 11** (FBA) for aquaculture probiotics
3. **Run Module 12** (CRISPR simulation) for aquaculture pathogens
4. **Design pilot trial** (e.g., tilapia + *A. hydrophila* + *B. subtilis* + phage)
5. **Proceed to Module 14** (System Biology Integration)

---

*Extension module for pr0117-PhageProbioticSynergy*
*Focus: Aquaculture-specific optimization*
