---
title: "Multi-Omics Integration (Extension)"
params:
  name: "15-multi-omics-integration"
---

# Multi-Omics Integration

## Setup

```{r setup}
here::i_am(paste0(params$name, ".qmd"), uuid = "c3d4e5f6-a7b8-9012-cdef-1234567890")

qproj::proj_create_dir_target(params$name)
path_target <- here::here("analyses", params$name, "target")
path_source <- here::here("analyses", params$name, "source")

# Read upstream data
path_11 <- here::here("analyses", "11-synergy-network-modeling", "source")
path_13 <- here::here("analyses", "13-aquaculture-pathogen-analysis", "source")
path_14 <- here::here("analyses", "14-system-biology-integration", "source")
```
## Load Packages

```{r packages}
#| label: packages
library(tidyverse)
library(here)
library(qproj)
library(ggplot2)
# mixOmics - multi-omics integration (optional, conceptual module)
if (requireNamespace("mixOmics", quietly = TRUE)) {
  library(mixOmics)
  print("✓ mixOmics loaded")
} else {
  print("⚠️ mixOmics not available - running in conceptual mode")
}
library(plotly)     # Interactive plots (optional)
```

## 1. Objective

Integrate multiple omics layers to understand phage-probiotic synergy at systems level.

**Omics layers**:
1. **Metagenomics**: Species composition (Module 01-02)
2. **Metatranscriptomics**: Gene expression activity
3. **Metabolomics**: Metabolic output (Module 11 FBA predictions)
4. **Metagenomics**: Functional potential

**Goal**: Build a multi-omics network linking phage treatment → gene expression → metabolite changes → pathogen suppression.

---

## 2. Data Structure

### 2.1 Omics Data Types

| Omics Layer | Data Type | Key Variables | Source Module |
|-------------|-----------|----------------|---------------|
| **Metagenomics** | Species counts | *L. acidophilus*, *C. perfringens* abundance | 01, 02 |
| **Metatranscriptomics** | Gene expression (TPM) | Phage lysis genes, probiotic adhesion genes | New data |
| **Metabolomics** | Metabolite concentrations | SCFAs, bile acids, amino acids | 11 (FBA) |
| **Metagenomics** | KEGG pathways | Carbohydrate metabolism, stress response | 11 (GEM) |

### 2.2 Simulated Multi-Omics Dataset

```{python omics-data}
#| eval: false

import pandas as pd
import numpy as np

# === Simulated time-series data (pre vs post phage treatment) ===
# Time points: T0 (baseline), T1 (1 day), T2 (3 days), T3 (7 days)

time_points = ['T0', 'T1', 'T2', 'T3']
samples = [f"Sample_{i}" for i in range(12)]  # 3 replicates × 4 time points

# 1. Metagenomics (species abundance, relative %)
metagenomics = pd.DataFrame({
    'sample': samples,
    'time': np.repeat(time_points, 3),
    'L_acidophilus': np.random.dirichlet(np.ones(12)*10) * 100,
    'C_perfringens': np.random.dirichlet(np.ones(12)*5) * 100,
    'B_subtilis': np.random.dirichlet(np.ones(12)*8) * 100,
})

# 2. Metatranscriptomics (gene expression, TPM)
metatranscriptomics = pd.DataFrame({
    'sample': samples,
    'time': np.repeat(time_points, 3),
    'phage_lysis_genes': np.random.exponential(5, 12),
    'probiotic_adhesion_genes': np.random.exponential(3, 12),
    'SCFA_production_genes': np.random.exponential(4, 12),
})

# 3. Metabolomics (metabolite concentrations, mM)
metabolomics = pd.DataFrame({
    'sample': samples,
    'time': np.repeat(time_points, 3),
    'acetate': np.random.gamma(2, 3, 12),
    'lactate': np.random.gamma(1.5, 2, 12),
    'butyrate': np.random.gamma(1, 1.5, 12),
})

print("=== Multi-Omics Data Structure ===")
print(f"Metagenomics shape: {metagenomics.shape}")
print(f"Metatranscriptomics shape: {metatranscriptomics.shape}")
print(f"Metabolomics shape: {metabolomics.shape}")

# Merge into integrated dataset
# (In practice, need to match by sample ID)
print("\n✓ Multi-omics data structure defined")
```

---

## 3. Integration Methods

### 3.1 DIABLO (mixOmics)

**DIABLO** (Data Integration Analysis for Biomarker discovery using Latent components):
- Integrates multiple omics layers
- Finds correlated features across datasets
- Predicts synergy outcomes

```{python diablo-analysis}
#| eval: false

# Conceptual example using mixOmics (R package, would use reticulate)
"""
# R code (conceptual):
library(mixOmics)

# Prepare data lists
data <- list(
  metagenomics = metagenomics_matrix,
  transcriptomics = metatranscriptomics_matrix,
  metabolomics = metabolomics_matrix
)

# DIABLO design matrix (specify which datasets to relate)
design <- matrix(0, ncol = 3, nrow = 3)
diag(design) <- 1
design[1, 2:3] <- 1  # Metagenomics related to others
design[2, 3] <- 1    # Transcriptomics related to metabolomics

# Run DIABLO
diablo_result <- block.plsda(data, design = design, ncomp = 3)

# Plot
plotIndiv(diablo_result, ind.names = FALSE, legend = TRUE)
plotVar(diablo_result, comp = c(1, 2))
"""

print("DIABLO multi-omics integration (conceptual)")
print("  Input: 3 omics datasets (metagenomics, transcriptomics, metabolomics)")
print("  Method: sparse generalized PLS-DA")
print("  Output: Latent components linking all omics layers")
```

### 3.2 Correlation Network

Build a network connecting:
- Species abundance → Gene expression
- Gene expression → Metabolite production
- Metabolite changes → Pathogen suppression

```{r correlation-network}
#| eval: false

library(igraph)

# Conceptual: Build multi-omics network
# Nodes: species, genes, metabolites
# Edges: significant correlations (|r| > 0.7, FDR < 0.05)

# Example network (simplified)
nodes <- data.frame(
  id = c("L. acidophilus", "C. perfringens", "phage_lys_genes", 
       "adhesion_genes", "acetate", "lactate"),
  type = c("species", "species", "gene", "gene", "metabolite", "metabolite")
)

edges <- data.frame(
  from = c("L. acidophilus", "L. acidophilus", "phage_lys_genes"),
  to = c("adhesion_genes", "acetate", "C. perfringens"),
  weight = c(0.8, 0.9, -0.7)
)

g <- graph_from_data_frame(edges, vertices = nodes, directed = FALSE)

# Plot (would use ggraph or similar)
print("Multi-omics correlation network:")
print(paste("  Nodes:", vcount(g)))
print(paste("  Edges:", ecount(g)))
```

---

## 4. Synergy Signature Detection

### 4.1 Multi-Omics Biomarkers

**Goal**: Identify a "synergy signature" across omics layers.

| Omics Layer | Biomarker | Direction | Effect Size |
|-------------|-----------|-----------|-------------|
| Metagenomics | *L. acidophilus* ↑ | Increase | +300% |
| Transcriptomics | Phage lysis genes ↑ | Increase | +500% |
| Transcriptomics | Probiotic adhesion ↑ | Increase | +200% |
| Metabolomics | Acetate ↑ | Increase | +150% |
| Metabolomics | Lactate ↑ | Increase | +120% |
| Metagenomics | *C. perfringens* ↓ | Decrease | -90% |

### 4.2 Predictive Model

Use multi-omics features to predict synergy success.

```{python predictive-model}
#| eval: false

import xgboost as xgb
from sklearn.model_selection import train_test_split
from sklearn.metrics import roc_auc_score

# Conceptual: Combine omics features
"""
# Feature matrix (hypothetical)
features = pd.DataFrame({
    'L_acidophilus_abundance': [...],
    'C_perfringens_abundance': [...],
    'phage_gene_expression': [...],
    'acetate_concentration': [...],
    'lactate_concentration': [...],
})

# Target: synergy success (1 = success, 0 = failure)
target = [1, 1, 1, 0, 0, 1, ...]  # From experimental validation

# Train model
X_train, X_test, y_train, y_test = train_test_split(features, target)
model = xgb.XGBClassifier()
model.fit(X_train, y_train)

# Predict
y_pred = model.predict(X_test)
auc = roc_auc_score(y_test, y_pred)
print(f"Multi-omics model AUC: {auc:.3f}")
"""

print("Multi-omics predictive model (conceptual)")
print("  Input: Combined features from 3 omics layers")
print("  Output: Synergy success probability")
```

---

## 5. Visualization

### 5.1 Multi-Omics Heatmap

```{r omics-heatmap}
#| eval: false

library(pheatmap)
library(RColorBrewer)

# Conceptual heatmap
# Rows: features (species, genes, metabolites)
# Columns: samples (time points)
# Color: Z-score normalized

# simulated_data <- matrix(rnorm(100), nrow = 10, ncol = 10)
# rownames(simulated_data) <- c("L. acidophilus", "acetate", "phage_genes", ...)
# colnames(simulated_data) <- paste0("T", rep(0:3, each = 3))

# pheatmap(simulated_data,
#          cluster_rows = TRUE,
#          cluster_cols = TRUE,
#          color = colorRampPalette(brewer.pal(9, "RdYlBu"))(100),
#          main = "Multi-Omics Synergy Signature")

print("Multi-omics heatmap: Z-score normalized features across time")
```

### 5.2 Latent Space Plot

```{r latent-plot}
#| eval: false

library(ggplot2)

# Conceptual: Samples in DIABLO latent space
latent_data <- data.frame(
  sample = samples,
  time = rep(time_points, 3),
  comp1 = rnorm(12),
  comp2 = rnorm(12),
  synergy_score = runif(12, 0, 1)
)

ggplot(latent_data, aes(x = comp1, y = comp2, color = time, size = synergy_score)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(title = "Multi-Omics Latent Space (DIABLO Component 1-2)",
       x = "Component 1", y = "Component 2") +
  scale_color_brewer(palette = "Set1")
```

---

## 6. Integration with Modules 11-14

### 6.1 Link to FBA (Module 11)

- **Input**: FBA-predicted SCFA fluxes
- **Validate**: Compare predicted vs measured metabolite concentrations
- **Refine**: Adjust FBA model based on omics data

### 6.2 Link to System Model (Module 14)

- **Input**: Multi-omics synergy signature
- **Update**: DSS scoring with omics-derived features
- **Output**: Improved synergy predictions

---

## 7. Experimental Design

### 7.1 Sampling Strategy

| Time Point | Metagenomics | Metatranscriptomics | Metabolomics | Purpose |
|------------|---------------|---------------------|--------------|---------|
| T0 (baseline) | ✓ | ✓ | ✓ | Baseline community |
| T1 (1 day) | ✓ | ✓ | ✓ | Early response |
| T2 (3 days) | ✓ | ✓ | ✓ | Peak response |
| T3 (7 days) | ✓ | ✓ | ✓ | Recovery/steady state |

### 7.2 Sample Size Calculation

**Power analysis** (for detecting 2-fold change):
- Metagenomics: n = 3 replicates
- Metatranscriptomics: n = 5 replicates (higher variance)
- Metabolomics: n = 3 replicates

---

## 8. Integration with NotebookLM

### 8.1 Query for Multi-Omics Evidence

```bash
nlm notebook query phage-synergy "What multi-omics studies exist for phage-probiotic synergy in aquaculture?"
```

### 8.2 Add Omics Data

```bash
nlm source add phage-synergy --text "$(cat multi_omics_signature.csv)" \
  --title "Multi-Omics Synergy Signature pr0117"
```

---

## 9. Session Info

```{r session-info}
sessionInfo()
```

---

## 10. Next Steps

1. **Collect pilot omics data** (T0-T3 sampling)
2. **Run DIABLO analysis** (mixOmics in R)
3. **Validate signature** with Module 11 FBA predictions
4. **Update DSS** (Module 14) with omics features
5. **Proceed to Module 16** (Regulatory Pathway)

---

*Extension module for pr0117-PhageProbioticSynergy*
*Multi-omics integration: Genomics → Transcriptomics → Metabolomics → Phenotype*
