---
title: "System Biology Integration (Extension)"
params:
  name: "14-system-biology-integration"
---

# System Biology Integration

## Setup

```{r setup}
here::i_am(paste0(params$name, ".qmd"), uuid = "c3d4e5f6-a7b8-9012-cdef-1234567890")

qproj::proj_create_dir_target(params$name)
path_target <- here::here("analyses", params$name, "target")
path_source <- here::here("analyses", params$name, "source")

# Read upstream data from all previous modules
path_03 <- here::here("analyses", "03-probiotic-target-identification", "source")
path_11 <- here::here("analyses", "11-synergy-network-modeling", "source")
path_12 <- here::here("analyses", "12-crispr-evolution-simulation", "source")
path_13 <- here::here("analyses", "13-aquaculture-pathogen-analysis", "source")
```

## Load Packages

```{r packages}
library(tidyverse)
library(here)
library(qproj)
library(ggplot2)
library(patchwork)  # For combining plots
library(cowplot)    # For publication-ready plots
```

## 1. Objective

Integrate all previous analyses into a **multi-scale system model** of phage-probiotic synergy.

**Integration levels**:
1. **Molecular**: Phage-host receptor interactions (Module 02)
2. **Metabolic**: FBA predictions (Module 11)
3. **Evolutionary**: CRISPR escape dynamics (Module 12)
4. **Ecological**: Aquaculture system context (Module 13)
5. **System-level**: Combined effect prediction

**Goal**: Provide a decision support system (DSS) for optimizing phage-probiotic therapy in aquaculture.

---

## 2. Multi-Scale Data Integration

### 2.1 Data Sources Summary

| Module | Data Type | Scale | Key Output |
|--------|-----------|-------|-------------|
| 02 | Host prediction | Molecular | Phage-host pairs |
| 03 | ML ranking | Strain | Synergy scores |
| 11 | FBA | Metabolic | Flux changes |
| 12 | Evolution | Population | Escape risk |
| 13 | Aquaculture | System | Environmental fit |

### 2.2 Integrated Data Frame

Combine insights from all modules.

```{python integrated-data}
#| eval: false

import pandas as pd
import numpy as np

# === Simulated integrated data (would come from actual module outputs) ===
integrated_data = pd.DataFrame({
    'probiotic': ['L_acidophilus', 'B_subtilis', 'B_lactis', 'E_faecium'],
    'target_pathogen': ['C_perfringens', 'C_perfringens', 'S_enterica', 'C_perfringens'],
    'phage_available': [True, True, False, True],
    'synergy_score': [0.82, 0.55, 0.78, 0.75],  # From Module 03
    'fba_growth_increase': [0.49, 0.32, 0.41, 0.38],  # From Module 11 (49%, etc.)
    'crispr_risk': ['Low', 'High', 'Medium', 'Low'],  # From Module 12
    'aquaculture_fit': ['Moderate', 'High', 'Moderate', 'Low'],  # From Module 13
    'env_stability': [0.7, 0.9, 0.75, 0.65]  # Environmental stability score
})

# Compute system-level synergy score (weighted)
# Weights: synergy (0.3), FBA (0.3), CRISPR risk (0.2, inverted), aquaculture (0.2)
def risk_score_to_numeric(risk):
    return {'Low': 1.0, 'Medium': 0.6, 'High': 0.2}[risk]

def fit_to_numeric(fit):
    return {'Low': 0.3, 'Moderate': 0.6, 'High': 1.0}[fit]

integrated_data['crispr_risk_numeric'] = integrated_data['crispr_risk'].apply(risk_score_to_numeric)
integrated_data['aquaculture_fit_numeric'] = integrated_data['aquaculture_fit'].apply(fit_to_numeric)

integrated_data['system_synergy_score'] = (
    integrated_data['synergy_score'] * 0.3 +
    integrated_data['fba_growth_increase'] * 0.3 +
    integrated_data['crispr_risk_numeric'] * 0.2 +
    integrated_data['aquaculture_fit_numeric'] * 0.2
)

# Sort by system-level score
integrated_sorted = integrated_data.sort_values('system_synergy_score', ascending=False)

print("=== Integrated System-Level Ranking ===")
print(integrated_sorted[['probiotic', 'target_pathogen', 'synergy_score', 
                       'fba_growth_increase', 'crispr_risk', 
                       'aquaculture_fit', 'system_synergy_score']])
```

---

## 3. Decision Support System (DSS) Framework

### 3.1 Scoring Algorithm

**Input parameters** (user-adjustable):
- `w_synergy`: Weight for ML synergy score (default 0.3)
- `w_fba`: Weight for FBA prediction (default 0.3)
- `w_crispr`: Weight for CRISPR risk (default 0.2)
- `w_aqua`: Weight for aquaculture fit (default 0.2)

**Output**: System-level recommendation with confidence interval.

### 3.2 Interactive Recommendation

```{python dss-engine}
#| eval: false

def compute_dss_score(row, w_synergy=0.3, w_fba=0.3, w_crispr=0.2, w_aqua=0.2):
    """
    Compute Decision Support System score.
    Higher = better candidate.
    """
    # Invert CRISPR risk (Low risk = high score)
    crispr_score = {'Low': 1.0, 'Medium': 0.6, 'High': 0.2}[row['crispr_risk']]
    
    # Aquaculture fit
    aqua_score = {'Low': 0.3, 'Moderate': 0.6, 'High': 1.0}[row['aquaculture_fit']]
    
    # Weighted sum
    dss_score = (
        row['synergy_score'] * w_synergy +
        row['fba_growth_increase'] * w_fba +
        crispr_score * w_crispr +
        aqua_score * w_aqua
    )
    
    return dss_score

# Apply DSS to integrated data
integrated_data['dss_score'] = integrated_data.apply(
    lambda row: compute_dss_score(row), axis=1
)

# Rank
dss_ranked = integrated_data.sort_values('dss_score', ascending=False)

print("=== DSS Recommendations ===")
print(dss_ranked[['probiotic', 'target_pathogen', 'dss_score']].to_string(index=False))

# Top recommendation
top = dss_ranked.iloc[0]
print(f"\n*** Top Recommendation ***")
print(f"Probiotic: {top['probiotic']}")
print(f"Target Pathogen: {top['target_pathogen']}")
print(f"DSS Score: {top['dss_score']:.3f}")
print(f"Rationale:")
print(f"  - Synergy Score: {top['synergy_score']:.2f}")
print(f"  - FBA Growth Increase: {top['fba_growth_increase']:.0%}")
print(f"  - CRISPR Risk: {top['crispr_risk']}")
print(f"  - Aquaculture Fit: {top['aquaculture_fit']}")
```

---

## 4. System Dynamics Visualization

### 4.1 Multi-Panel Plot

Combine key insights into a single figure.

```{r system-viz}
#| eval: false

# Load integrated data (in practice, read from Python output)
# integrated_df <- read_csv("integrated_data.csv")

# Create example data for visualization
system_data <- tibble(
  probiotic = rep(c("L. acidophilus", "B. subtilis", "B. lactis", "E. faecium"), 2),
  metric = rep(c("Synergy Score", "System DSS Score"), each = 4),
  value = c(0.82, 0.55, 0.78, 0.75,  # Synergy scores
            0.71, 0.62, 0.68, 0.65)  # DSS scores (example)
)

# Plot
p <- ggplot(system_data, aes(x = probiotic, y = value, fill = metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Probiotic Candidates: Synergy vs System DSS Score",
       x = "", y = "Score") +
  scale_fill_brewer(palette = "Set2") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p)
```

### 4.2 Network-Genome-FBA Integration

**Conceptual diagram** (to be created with Cytoscape or similar):
- **Layer 1**: Molecular network (phage-pathogen-probiotic)
- **Layer 2**: Metabolic network (FBA reactions)
- **Layer 3**: Evolutionary dynamics (CRISPR escape)
- **Layer 4**: System context (aquaculture environment)

---

## 5. Uncertainty Quantification

### 5.1 Sensitivity Analysis

Test how DSS scores change with different weight assignments.

```{python sensitivity-analysis}
#| eval: false

import itertools

# Define weight ranges
weight_ranges = {
    'w_synergy': [0.2, 0.3, 0.4],
    'w_fba': [0.2, 0.3, 0.4],
    'w_crispr': [0.1, 0.2, 0.3],
    'w_aqua': [0.1, 0.2, 0.3]
}

# Generate all combinations (simplified: only test a few)
combinations = list(itertools.product(*weight_ranges.values()))
print(f"Total combinations: {len(combinations)}")

# Test top 3 combinations
top_combinations = combinations[:3]

for i, combo in enumerate(top_combinations):
    w_syn, w_fba, w_crispr, w_aqua = combo
    # Normalize weights
    total = w_syn + w_fba + w_crispr + w_aqua
    w_syn, w_fba, w_crispr, w_aqua = w_syn/total, w_fba/total, w_crispr/total, w_aqua/total
    
    # Compute scores
    scores = []
    for _, row in integrated_data.iterrows():
        score = compute_dss_score(row, w_syn, w_fba, w_crispr, w_aqua)
        scores.append(score)
    
    integrated_data[f'dss_combo_{i}'] = scores
    
    print(f"\nCombo {i+1}: synergy={w_syn:.1f}, fba={w_fba:.1f}, crispr={w_crispr:.1f}, aqua={w_aqua:.1f}")
    print(f"  Top DSS score: {max(scores):.3f}")

# Check ranking stability
print("\n=== Ranking Stability ===")
for i in range(3):
    ranked = integrated_data.sort_values(f'dss_combo_{i}', ascending=False)
    print(f"Combo {i+1} top 3: {list(ranked['probiotic'].head(3))}")
```

---

## 6. Implementation Roadmap

### 6.1 Phased Deployment

| Phase | Action | Timeline | Modules Integrated |
|-------|--------|----------|-------------------|
| **Phase 1** | Pilot trial (top candidate) | Month 1-3 | 03, 11 |
| **Phase 2** | Add aquaculture optimization | Month 4-6 | 13, 14 |
| **Phase 3** | Monitor CRISPR evolution | Month 6-12 | 12, 14 |
| **Phase 4** | Scale to commercial | Month 12+ | All |

### 6.2 Success Metrics

- **Primary**: Pathogen load reduction >90% (qPCR)
- **Secondary**: Probiotic colonization >10⁶ CFU/g gut
- **Tertiary**: No CRISPR escape mutants (monitoring)
- **Economic**: ROI >200% (cost of phage + probiotic vs losses saved)

---

## 7. Integration with NotebookLM

### 7.1 Query for System-Level Evidence

```bash
nlm notebook query phage-synergy "What are the best practices for integrating multi-omics data in probiotic therapy?"
```

### 7.2 Add System Model

```bash
nlm source add phage-synergy --text "$(cat integrated_dss_scores.csv)" \
  --title "System DSS Scores pr0117"
```

---

## 8. Session Info

```{r session-info}
sessionInfo()
```

---

## 9. Next Steps

1. **Validate DSS** with experimental data (Module 07-08)
2. **Refine weights** based on pilot trial results
3. **Expand to more probiotics** (Module 03 update)
4. **Add economic modeling** (Module 10 integration)
5. **Prepare manuscript** (Module 08 - incorporate system model)

---

*Extension module for pr0117-PhageProbioticSynergy*
*Multi-scale integration: Molecular → Metabolic → Evolutionary → System*
